---
title: "Doublet QC and downsampling"
author: "George Howitt"
date: "2024-09-19"
output: html_document
---

Looking at how doublet removal affects clustering. The 10X PBMC data set has been clustered at resolutions between 0.1 and 1. 

```{r setup}
suppressPackageStartupMessages({
  library(BiocStyle)
  library(BiocParallel)
  library(tidyverse)
  library(here)
  library(glue)
  library(janitor)
  library(cowplot)
  library(patchwork)
  library(scales)
  library(readr)
  library(Seurat)
  library(Matrix)
})
```

# Data loading

Load the PBMC data with the doublets identified but not removed and with doublets removed.

```{r}
seu <- readRDS(here("data", "10X_PBMC_16plex", "SEU",
             "10X_PBMC.CellRanger.decontX.mito.doublet.filter.Azimuth.SEU.rds"))

seu_doublets <- readRDS(here("data", "10X_PBMC_16plex", "SEU",
             "10X_PBMC.CellRanger.decontX.mito.filter.Azimuth.SEU.rds"))
```

# Is clustering driven by doublets?

Number of clusters vs resolution

```{r}
cluster_resolutions <- c(seq(0.1,0.9,0.1),1)
cluster_df <- data.frame(resolution = cluster_resolutions,
                 No = sapply(cluster_resolutions, function(i) length(levels(unlist(seu[[paste0("RNA_snn_res.",i)]])))),
                 Yes = sapply(cluster_resolutions, function(i) length(levels(unlist(seu_doublets[[paste0("RNA_snn_res.",i)]]))))) %>%
  pivot_longer(cols = c(No, Yes), names_to = 'Doublets included?') 
```

```{r}
pa <- ggplot(cluster_df) + 
  geom_point(aes(x = resolution, y = value, color = `Doublets included?`, shape = `Doublets included?`), size = 2) + 
  xlab("Clustering resolution") + 
  ylab("N clusters") +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
panel.background = element_blank(), axis.line = element_line(colour = "black"),
legend.text = element_text(size = 6),
        legend.title = element_text(size = 8))
pa
```

```{r}
library(clustree)
clustree(seu_doublets)
clustree(seu)
```

Choosing resolution of 0.4 going forward

```{r}
pb <- ggplot(seu_doublets@meta.data, aes(x = RNA_snn_res.0.4, fill = scDblFinder.class)) + 
  xlab("Cluster") +
  geom_bar() + NoLegend() + 
  theme(axis.text.x = element_text(size=4))

pc <- ggplot(seu_doublets@meta.data, aes(x = RNA_snn_res.0.4, fill = scDblFinder.class)) + 
  xlab("Cluster") +
  geom_bar(position = 'fill') + 
  theme(axis.text.x = element_text(size=4),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 8))
  
pb | pc
```

How many clusters are dominated by doublets?
```{r}
doublet_fraction <- sum(seu_doublets$scDblFinder.class == 'doublet')/ncol(seu_doublets)
t1 <- tabyl(seu_doublets@meta.data, RNA_snn_res.0.4, scDblFinder.class) %>% adorn_percentages()
sum(t1$doublet > 2 * doublet_fraction)
```

How many clusters have more than twice the average fraction of doublets?
```{r}
cluster_doublet_fractions <- sapply(levels(seu_doublets$RNA_snn_res.0.4),
                                    function(cluster) sum(seu_doublets$RNA_snn_res.0.4 == cluster & seu_doublets$scDblFinder.class == 'doublet') /
                                      sum(seu_doublets$RNA_snn_res.0.4 == cluster))

names(cluster_doublet_fractions > 2*sum(seu_doublets$scDblFinder.class=='doublet')/ncol(seu_doublets))[cluster_doublet_fractions > 2*sum(seu_doublets$scDblFinder.class=='doublet')/ncol(seu_doublets)] 
```

```{r}
p1 <- DimPlot(seu_doublets, group.by = "predicted.celltype.l2", raster=F, label = T, label.size = 2,  repel = T) +
  ggtitle("Doublets included")

p2 <- DimPlot(seu_doublets[, seu_doublets$RNA_snn_res.0.4 %in% c("12", "14", "15", "16", "18", "19", "21", "22", "25", "27", "28")], group.by = "predicted.celltype.l2", raster=F, label = T, label.size = 2, repel = T) +
  xlim(-15,15) +
  ylim(-15,15) +
  ggtitle("Doublet clusters")

p3 <- DimPlot(seu, group.by = "predicted.celltype.l2", label = T, label.size = 2, repel = T) + 
  ggtitle("Singlets only") 

pdef <- (p1 | p2 | p3) & 
  NoLegend() & 
  theme(axis.text = element_blank(), axis.title = element_blank(), axis.ticks =  element_blank())
```

```{r}
((pa | pb | pc) / pdef) +  plot_annotation(tag_levels = 'a')
```

# Downsampling experiment

Randomly subsample doublet object to make it same size as singlets only object. This step takes a long time. 
```{r}
for(i in 1:10) {
  seu_downsample <- seu_doublets[, sample.int(n = ncol(seu_doublets), size = ncol(seu))]
  seu_downsample <- NormalizeData(seu_downsample, verbose = F) %>%
  FindVariableFeatures(nfeatures = 2000, verbose = F) %>%
  ScaleData(verbose = F) %>%
  RunPCA(dims = 1:30, verbose = F) %>%
  RunUMAP(dims = 1:30, verbose = F) %>% 
  FindNeighbors(reduction = "pca", dims = 1:30) %>%
  FindClusters(resolution = seq(0.1, 1, by = 0.1), algorithm = 3)
  saveRDS(seu_downsample, here("data", "10X_PBMC_16plex", "SEU",
             paste0("10X_PBMC.CellRanger.decontX.mito.filter.Azimuth.downsample", as.character(i),".SEU.rds")))
}
```

```{r}
p1 <- ggplot(seu_downsample@meta.data, aes(x = RNA_snn_res.0.4, fill = scDblFinder.class)) + 
  geom_bar() + NoLegend()

p2 <- ggplot(seu_downsample@meta.data, aes(x = RNA_snn_res.0.4, fill = scDblFinder.class)) + 
  geom_bar(position = 'fill') 
  
(p1 | p2) &
  xlab("Cluster")
```

Make a dataframe to store the summary level information from each of the downsampled runs
```{r}
cluster_res <- c(seq(0.1,0.9,0.1),1)
n_clusters = NULL
doublet_fraction = NULL
mean_doublet_fraction = sum(seu_doublets$scDblFinder.class == 'doublet') / ncol(seu_doublets)

for (i in 1:10) {
  seu_downsample <- readRDS(here("data", "10X_PBMC_16plex", "SEU",
             paste0("10X_PBMC.CellRanger.decontX.mito.filter.Azimuth.downsample", as.character(i),".SEU.rds")))
  n_clusters = cbind(n_clusters, sapply(cluster_res, function(i) length(levels(unlist(seu_downsample[[paste0("RNA_snn_res.",i)]])))))
  
  doublet_fraction <- c(doublet_fraction, 
                        sum(sapply(unique(seu_downsample$RNA_snn_res.0.4), function(c) sum(seu_downsample$RNA_snn_res.0.4 == c & seu_downsample$scDblFinder.class == 'doublet') / sum(seu_downsample$RNA_snn_res.0.4 == c)) > 2*mean_doublet_fraction))
}
```

```{r}
temp <- n_clusters
```

```{r}
mean_clusters <- apply(temp, 1, mean)
sd_cluster <- apply(temp, 1, sd)
downsample_df <- data.frame(resolution = cluster_resolutions,
                            mean = mean_clusters,
                            ymin = mean_clusters - sd_cluster,
                            ymax = mean_clusters + sd_cluster)

```

```{r}
cluster_resolutions <- c(seq(0.1,0.9,0.1),1)
cluster_df <- data.frame(resolution = cluster_resolutions,
                 Singlets = sapply(cluster_resolutions, function(i) length(levels(unlist(seu[[paste0("RNA_snn_res.",i)]])))),
                 Doublets = sapply(cluster_resolutions, function(i) length(levels(unlist(seu_doublets[[paste0("RNA_snn_res.",i)]])))),
                 Downsampled = mean_clusters) %>%
  pivot_longer(cols = c(Singlets, Doublets, Downsampled)) 
```

```{r}
ggplot(cluster_df) +
  geom_point(aes(x = resolution, y = value, color = name, shape = name)) +
  xlab('Cluster resoltuion') +
  ylab('N clusters') +
  theme(legend.title = element_blank())
  #geom_errorbar(data = downsample_df, aes(x = resolution, y = mean, ymin = ymin, ymax = ymax))
```

```{r}
temp <- as.data.frame(temp)
rownames(temp) <- seq(0.1,1,0.1)
colnames(temp) <- paste0("Run ", as.character(1:10))
temp <- temp %>%
  pivot_longer(cols = 1:10)
```



