---
title: "Modelling and simulating doublets"
author: "George Howitt"
date: "2025-04-08"
output: html_document
---

Figured out that the model for predicting doublets in the original version of the paper was actually wrong. Going through the details here. 

```{r}
library(partitions)
library(ggplot2)
library(tidyverse)
library(matrixStats)
```

# Simulating doublets

Load in the data from the paper
```{r}
GEM_data <- readRDS("~/OneDrive - Peter Mac/singlecell/flex_multiplets/data/16plex_summary_data.rds")
```

Using inverse transform sampling to get pseudo-random numbers from a zero-truncated Poisson distribution.
Borrowing some code from here https://rpubs.com/binhho660/902552
```{r}
discrete_inverse_sampling <- function( U, prob ) {
  if(U <= prob[1]){
    return(1)
  }
  for(state in 2:length(prob)) {
    if(sum(prob[1:(state-1)]) < U && U <= sum(prob[1:state]) ) {
      return(state)
    }
  }
}
```

Zero-truncated Poisson distribution
```{r}
ztPois <- function(k, X) {
  lambda = uniroot(function(x) x / (1 - exp(-x)) - X, interval = c(0.1,5))$root
  
  return(lambda^k / ((exp(lambda) - 1) * factorial(k)))
}
```

This function gives a doublet fraction based on an input mean droplet occupancy, k. 
```{r}
simulate_doublet_fraction <- function(X, n = 16, i = 10000, prob = NULL) {
  
  #Generate the droplet occupancy
  k <- sapply(runif(i), function(u) discrete_inverse_sampling(u, ztPois(1:16, X)))
  
  #Generate the "cells" in each "droplet"
  BCs <- 1:n
  cell_ids <- sapply(k, function(x) sample(BCs, x, replace = T, prob = prob))
  
  #Number of doublets is the number of barcodes with counts > 1 in each droplet
  ndoublets <- sapply(cell_ids, function(d) sum(table(d)>1))
  
  ncells <- sum(sapply(cell_ids, function(d) length(d)))
  
  #Generate the apparent cells
  unique_BCs <- sapply(cell_ids, function(d) unique(d))
  ncells_observed <- sum(sapply(unique_BCs, function(d) length(d)))
  ndoublets <- sum(ndoublets)
    
  #Compute the doublet fraction
  return(ndoublets/ncells_observed)
}
```

And this uses the same concept to give a value for the observed mean occupancy.
Can also simulate the distribution of apparent occupancies from the true occupancy distribution. 
```{r}
simulate_apparent_occupancy <- function(X, n = 16, i = 10000) {
  
  #Generate the droplet occupancy
  k <- sapply(runif(i), function(u) discrete_inverse_sampling(u, ztPois(1:16, X)))
  
  #Generate the "cells" in each "droplet"
  BCs <- 1:n
  cell_ids <- sapply(k, function(x) sample(BCs, x, replace = T))
  
  #Apparent size of droplets
  l <- sapply(cell_ids, function(d) length(unique(d)))
  
  return(mean(l))
}
```

Simulate apparent occupancies and doublet fractions on a grid of mean occupancies
```{r}
x <- seq(1.1,2.5,0.1)
l <- sapply(x, function(x) simulate_apparent_occupancy(x))
sim_frac <- sapply(x, function(x) simulate_doublet_fraction(x))
```

# Doublet fraction model - combinations

This is the model based on the model of combinatorics we put in the paper. It is not correct :(

```{r}
multiplet_configs <- function(n, k) {
  M_k = factorial(n + k - 1) / (factorial(n - 1) * factorial(k))
  return(M_k)
}

partition_configs_original <- function(n, k) {
  out = NULL #List that gets output by the function
  p = parts(k) #Matrix of partitions of k
  
  for (i in 1:ncol(p)) { 
    part = p[,i]
    part = part[part != 0] #Remove zeros
    u = unique(part) #Identify unique entries in the partition
    
    #Dummy variables
    counter = n #Keeps track of how many unique samples are left to choose from
    final = 1 # Output for the below loop (we are computing a product)
    
    for (j in u) { #j is a counter of the unique integers in the partition
      l = sum(part == j) #Number of entries of j in the partition
      combos = choose(counter, l) #Number of non-repeating combinations of choosing j
      counter = counter - l #Decrement n by l each time since there are now fewer unique samples remaining
      final = final*combos
    }
    out = c(out,final)
  }
  return(out)
}

doublet_fraction_observed_original <- function(n, k) {
  out = NULL #output of the function
  part_configs = partition_configs_original(n, k)
  p = parts(k)
  
  for (i in 1:ncol(p)) { # Go through each partition one by one
    part = p[,i]
    part = part[part != 0] #Remove zeros
    resolvable_cells = sum(part == 1)
    apparent_length = length(part)
    f = 1 - resolvable_cells / apparent_length #Fraction of observed cells that are resolvable
    doublet_fraction = f*part_configs[i]   
    out = c(out, doublet_fraction)
  }
  return(sum(out)/multiplet_configs(n,k))
}

doublet_frac_model_original <- function(X, n = 16) {
  return(sum(sapply(1:16, function(k) k*ztPois(k, X) * (doublet_fraction_observed_original(n,k)))) / 
           sum(sapply(1:16, function(k) k*ztPois(k, X))))
}
```

Version of the model from the paper. 
```{r}
multiplet_configs_combinations <- function(n, k) {
  M_k = factorial(n + k - 1) / (factorial(n - 1) * factorial(k))
  return(M_k)
}

partition_configs_combinations <- function(n, k, p ) {
  part = p[p != 0] #Remove zeros
  u = unique(p) #Identify unique entries in the partition
  
  #Dummy variables
  counter = n #Keeps track of how many unique samples are left to choose from
  out = 1 # Output for the below loop (we are computing a product)
  
  for (j in u) { #j is a counter of the unique integers in the partition
    l = sum(part == j) #Number of entries of j in the partition
    combos = choose(counter, l) #Number of non-repeating combinations of choosing j
    counter = counter - l #Decrement n by l each time since there are now fewer unique samples remaining
    out = out*combos
  }
  return(out)
}

doublet_fraction_combinations <- function(n, k) {
  p = parts(k)
  doublet_fractions_parts = sapply(p, function(part) sum(part > 1) / sum(part != 0))
  ncells_observed = sum(apply(p, 2, function(part) partition_configs_combinations(n, k, part) * sum(part != 0)))
  doublet_fraction = sum(apply(p, 2, function(part)  partition_configs_combinations(n, k, part) * sum(part > 1))) / ncells_observed
  return(doublet_fraction)
}

doublet_frac_combinations_distribution <- function(X, n = 16) {
  return(sum(sapply(1:16, function(k) k*ztPois(k, X) * (doublet_fraction_combinations(n,k)))) / 
           sum(sapply(1:16, function(k) k*ztPois(k, X))))
}
```

# Doublet fraction model - permutations
```{r}
partition_configs <- function(n, k, p) {
  
  part = p[p != 0] #Remove zeros
  multiset_permutations = factorial(k) / product(factorial(part))
  u = unique(part) #Identify unique entries in the partition
  
  #Dummy variables
  counter = n #Keeps track of how many unique samples are left to choose from
  final = 1 # Output for the below loop (we are computing a product)
  
  for (j in u) { #j is a counter of the unique integers in the partition
    l = sum(part == j) #Number of entries of j in the partition
    combos = choose(counter, l) #Number of non-repeating combinations of choosing j
    counter = counter - l #Decrement n by l each time since there are now fewer unique samples remaining
    final = final*combos
  }
  
  out = final*multiset_permutations
  return(out)
}

doublet_fraction_k <- function(n, k) {
  p = parts(k)
  doublet_fractions_parts = sapply(p, function(part) sum(part > 1) / sum(part != 0))
  ncells_observed = sum(apply(p, 2, function(part) partition_configs(n, k, part) * sum(part != 0)))
  doublet_fraction = sum(apply(p, 2, function(part)  partition_configs(n, k, part) * sum(part > 1))) / ncells_observed
  return(doublet_fraction)
}

doublet_frac_permutations_model <- function(X, n = 16) {
  return(sum(sapply(1:n, function(k) k*ztPois(k, X) * (doublet_fraction_k(n,k)))) / 
           sum(sapply(1:n, function(k) k*ztPois(k, X))))
}
```

# Comparison
Make a plot to compare them all
```{r}
doublet_fraction_df <- data.frame("X" = x,
                                  "l" = l,
                                  "Simulation" = sim_frac,
                                  "Original" = sapply(x, function(x) doublet_frac_model_original(x)),
                                  "Combinations" = sapply(x, function(x) doublet_frac_combinations_distribution(x)),
                                  "Permutations" = sapply(x, function(x) doublet_frac_permutations_model(x))) %>%
  pivot_longer(cols = 3:6, names_to = "Model", values_to = "Doublet fraction")
doublet_fraction_df
```

```{r}

exp_data <- GEM_data %>%
  dplyr::select(c("mean_occupancy", "doublets_10X")) %>%
  rename(X = "mean_occupancy", `Doublet fraction` = "doublets_10X")

p1 <- ggplot(doublet_fraction_df[doublet_fraction_df$Model %in% c("Original", "Simulation"),]) +
  geom_point(mapping = aes(x = X,
                           y = `Doublet fraction`,
                           shape = Model, color = Model)) +
  xlab("Mean occupancy <k>")

p2 <- ggplot(doublet_fraction_df) +
  geom_point(mapping = aes(x = l,
                           y = `Doublet fraction`,
                           shape = Model, color = Model)) +
  geom_point(data = exp_data, aes(x = X,
                           y = `Doublet fraction`)) +
  xlab("Apparent mean occupancy <l>")

p1 
p2
```

```{r}
p1 <- ggplot(doublet_fraction_df[doublet_fraction_df$Model %in% c("Original", "Simulation"),]) +
  geom_point(mapping = aes(x = X,
                           y = `Doublet fraction`,
                           shape = Model, color = Model)) +
  #geom_point(data = exp_data, aes(x = X,
  #                         y = `Doublet fraction`)) +
  xlab("Mean occupancy <k>") 
p1
```

```{r}
flex_data_df <- data.frame('mean_occupancy' = GEM_data$mean_occupancy,
                           'Documentation' = GEM_data$doublets_10X,
                           'scDblFinder' = GEM_data$doublets_scDblFinder) %>%
  pivot_longer(2:3, names_to = 'Estimate from data', values_to = 'doublet fraction')
```


```{r}
p1 <- ggplot(doublet_fraction_df[doublet_fraction_df$Model %in% c("Original", "Simulation", "Permutations"),]) +
  geom_point(mapping = aes(x = X,
                           y = `Doublet fraction`,
                           color = Model), pch = 15) +
  geom_point(data = flex_data_df, aes(x = mean_occupancy,
                           y = `doublet fraction`, shape = `Estimate from data`)) + 

  xlab("Mean occupancy <k>") 
p1
```







